Properties and Bayesian fitting of restricted Boltzmann machines

### 2018-09-02 writing/paper-rbm/chapters/rbm.Rmd
intro wording tweaks

First, the models can be unsatisfactory as conceptualizations of how data are generated. That is, recalling @fisher1922mathematical, the aim of a statistical model is to represent data in a compact way. Neyman and Box further state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model; @box1967discrimination]. At issue, simulation from RBMs can often produce data lacking realistic variability so that such models may thereby fail to satisfactorily reflect a true data generation process. Such behavior relates to model degeneracy (or near degeneracy), which is a statistical concern in that most data processes being modeled are realistically not degenerate in their spectrum of potential outcomes. For example, when sampling data from a nearly degenerate RBM, only a small number of output possibilities have probability greater than zero, and thus a sample of images will all be copies of the same one image (or small number of images). An example of ten 4-pixel images simulated from a nearly degenerate RBM model is compared to ten 4-pixel images simulated from a non-degenerate RBM model in Figure \@ref(fig:sample-models). The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more potentially realistic variation. \par

@@

Separate of the exploration of model properties, we also investigate the quality of estimation in fitting these models and propose an estimation procedure that will allow for uncertainty quantification rather than point estimates alone. The fitting of RBMs can be problematic for two reasons, the first being computational and the second concerning flexibility. As the size of these models grows, both maximum likelihood and Bayesian methods of fitting quickly become intractable. The literature often suggests Markov chain Monte Carlo (MCMC) tools for approximate maximization of likelihoods to fit these models (e.g., Gibbs sampling to exploit conditional structure in hidden and visible variables), but little is said about the attributes of realized estimates [@hinton2010practical; @hinton2006fast], including the absence of interval estimates.

@@

While the computational concerns are inconvenient in fitting, issues due to model flexibility are potentially more concerning. For a RBM model with enough hidden variables, it has been shown that any distribution for the visibles can be approximated arbitrarily well [@le2008representational; @montufar2011refinements; and @montufar2011expressive]. However, for binary data, the empirical distribution of an observed training set of visible variables provides a highest likelihood benchmark, before any parametric model class is even introduced and applied to obtain a refinement of model fit. As a consequence, we find that any fully principled fitting method based on the likelihood for a RBM with enough hidden variables will seek to reproduce the (discrete) empirical distribution of a training set. This aspect can be undesirable, and perhaps even unexpected compared to most modeling scenarios, in that no "smoothed distribution" may result when fitting a RBM model of sufficient size with a rigorous likelihood-based method. We are therefore led to be skeptical that models that involve these structures (like deep Boltzmann machines) can achieve useful prediction or inference in a principled way without intentionally limiting the flexibility of the fitted model.

Notions of weight-decay (penalization) and sparsity (regularization) have been suggested in the RBM literature as practical remedies for both over-fitting and poor mixing in the Markov chain during fitting [@hinton2010practical; @tieleman2008training; @cho2012tikhonov]. Both $L_1$ and $L_2$ type penalties are mentioned to achieve weight-decay, though the benefits of these particular forms are unknown in different situations. However, the degree to which regularization and penalization are used by practitioners is not clear because these concepts are not (by default) a part of the accepted fitting methodology [@hinton2002training; @carreira2005contrastive]. In this paper, we attempt to address the concerns with overfitting and poor mixing in an alternative (and perhaps more transparent) manner, via specification of a Bayesian prior in Section \@ref(bayesian-model-fitting).

This paper is structured as follows. Section \@ref(background) formally defines the RBM including the joint distribution of hidden and visible variables and explains the model's connection to deep learning. Additionally, measures of model impropriety and methods of quantifying/detecting it are defined. Section \@ref(explorations-of-model-properties-through-simulation) details our explorations into the model behavior and potential propriety issues with the RBM class. This follows with the call to action in @sculley2018winner to employ small toy examples to advance empirical understanding of deep learning models and strive for a "higher level of empirical rigor in the field." We examine three Bayesian fitting techniques intended to avoid model impropriety issues raised in Section \@ref(model-fitting) and conclude with a discussion in Section \@ref(discussion). Supplementary online material provides proofs for results on RBM parameterizations and data codings described in Section \@ref(data-encoding).

>>>>

First, the models can be unsatisfactory as conceptualizations of how data are generated. That is, recalling @fisher1922mathematical, the aim of a statistical model is to represent data in a compact way. Neyman and Box further state that a model should "provide an explanation of the mechanism underlying the observed phenomena" [@lehmann1990model; @box1967discrimination]. At issue, simulation from RBMs can often produce data lacking realistic variability so that such models may thereby fail to satisfactorily reflect a true data generation process. Such behavior relates to model degeneracy (or near degeneracy), which is a statistical concern in that most data processes being modeled are realistically not degenerate in their spectrum of potential outcomes. For example, when sampling data from a nearly degenerate RBM, only a small number of output possibilities receives nearly all probability, and thus a sample of images will all be copies of the same one image (or small number of images). An example of ten 4-pixel images simulated from a nearly degenerate RBM model is compared to ten 4-pixel images simulated from a non-degenerate RBM model in Figure \@ref(fig:sample-models). The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more potentially realistic variation. \par

@@

Separate of the exploration of model properties, we also investigate the quality of estimation in fitting these models and examine an estimation procedure that allows for uncertainty quantification rather than point estimates alone. The fitting of RBMs can be problematic for two reasons, the first being computational and the second concerning flexibility. As the size of these models grows, both maximum likelihood and Bayesian methods of fitting quickly become intractable. The literature often suggests Markov chain Monte Carlo (MCMC) tools for approximate maximization of likelihoods to fit these models (e.g., Gibbs sampling to exploit conditional structure in hidden and visible variables), but little is said about the attributes of realized estimates [@hinton2010practical; @hinton2006fast], including the absence of interval estimates.

@@

As a compounding factor to the computational challenges, issues in model fitting due to model flexibility are potentially more concerning. For a RBM model with enough hidden variables, it has been shown that any distribution for the visibles can be approximated arbitrarily well [@le2008representational; @montufar2011refinements; and @montufar2011expressive]. However, for binary data, the empirical distribution of an observed training set of visible variables provides a highest likelihood benchmark, before any parametric model class is even introduced and applied to obtain a refinement of model fit. As a consequence, we find that any fully principled fitting method based on the likelihood for a RBM with enough hidden variables will seek to reproduce the (discrete) empirical distribution of a training set. This aspect can be undesirable, and perhaps even unexpected compared to most modeling scenarios, in that no "smoothed distribution" may result when fitting a RBM model of sufficient size with a rigorous likelihood-based method. We are therefore led to be skeptical that models that involve these structures (like deep Boltzmann machines) can achieve useful prediction or inference in a principled way without intentionally limiting the flexibility of the fitted model.

Notions of weight-decay (penalization) and sparsity (regularization) have been suggested in the RBM literature as practical remedies for both over-fitting and poor mixing in the Markov chain during fitting [@hinton2010practical; @tieleman2008training; @cho2012tikhonov]. Both $L_1$ and $L_2$ type penalties are mentioned to achieve weight-decay, though the benefits of these particular forms are unknown in different situations. However, the degree to which regularization and penalization are used by practitioners is not clear because these concepts are not (by default) a part of the overall model-specification-and-fitting methodology [@hinton2002training; @carreira2005contrastive]. In this paper, we attempt to address the concerns with overfitting and poor mixing in an alternative (and perhaps more transparent) manner, via specification of a Bayesian prior in Section \@ref(bayesian-model-fitting).

This paper is structured as follows. Section \@ref(background) formally defines the RBM including the joint distribution of hidden and visible variables and explains the model's connection to deep learning. Additionally, measures of model impropriety and methods of quantifying/detecting it are defined. Section \@ref(explorations-of-model-properties-through-simulation) details our explorations into the model behavior and potential propriety issues with the RBM class. This aligns with the call to action in @sculley2018winner to employ small toy examples to advance empirical understanding of deep learning models and strive for a "higher level of empirical rigor in the field." We examine three Bayesian fitting techniques intended to avoid model impropriety issues raised in Section \@ref(model-fitting) and conclude with a discussion in Section \@ref(discussion). Supplementary online material provides proofs for results on RBM parameterizations and data codings described in Section \@ref(data-encoding).

### 2018-09-02 writing/paper-rbm/docs/paper.tex
intro wording tweaks

\maketitle
\begin{abstract}
A restricted Boltzmann machine (RBM) is an undirected graphical model
constructed for discrete or continuous random variables, with two
layers, one hidden and one visible, and no conditional dependency within
a layer. In recent years, RBMs have risen to prominence due to their
connection to deep learning. By treating a hidden layer of one RBM as
the visible layer in a second RBM, a deep architecture can be created.
RBMs are thought to thereby have the ability to encode very complex and
rich structures in data, making them attractive for supervised learning.
However, the generative behavior of RBMs is largely unexplored and
typical fitting methodology does not easily allow for uncertainty
quantification in addition to point estimates. In this paper, we discuss
the relationship between RBM parameter specification in the binary case
and model properties such as degeneracy, instability and
uninterpretability. We also describe the difficulties that arise in
likelihood-based and discuss potential Bayes fitting of such (highly
flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are
often advocated for the RBM model structure.
\end{abstract}
\noindent%
{\it Keywords:}  Degeneracy, Instability, Classification, Deep Learning, Graphical Models
\vfill

@@

First, the models can be unsatisfactory as conceptualizations of how
data are generated. That is, recalling Fisher
(\protect\hyperlink{ref-fisher1922mathematical}{1922}), the aim of a
statistical model is to represent data in a compact way. Neyman and Box
further state that a model should ``provide an explanation of the
mechanism underlying the observed phenomena'' (Lehmann
\protect\hyperlink{ref-lehmann1990model}{1990}; G. E. P. Box
\protect\hyperlink{ref-box1967discrimination}{1967}). At issue,
simulation from RBMs can often produce data lacking realistic
variability so that such models may thereby fail to satisfactorily
reflect a true data generation process. Such behavior relates to model
degeneracy (or near degeneracy), which is a statistical concern in that
most data processes being modeled are realistically not degenerate in
their spectrum of potential outcomes. For example, when sampling data
from a nearly degenerate RBM, only a small number of output
possibilities have probability greater than zero, and thus a sample of
images will all be copies of the same one image (or small number of
images). An example of ten 4-pixel images simulated from a nearly
degenerate RBM model is compared to ten 4-pixel images simulated from a
non-degenerate RBM model in Figure \ref{fig:sample-models}. The
degenerate model places almost all probability on one outcome, causing
the image to be generated repeatedly, whereas the non-degenerate model
shows more potentially realistic variation.

@@

Separate of the exploration of model properties, we also investigate the
quality of estimation in fitting these models and propose an estimation
procedure that will allow for uncertainty quantification rather than
point estimates alone. The fitting of RBMs can be problematic for two
reasons, the first being computational and the second concerning
flexibility. As the size of these models grows, both maximum likelihood
and Bayesian methods of fitting quickly become intractable. The
literature often suggests Markov chain Monte Carlo (MCMC) tools for
approximate maximization of likelihoods to fit these models (e.g., Gibbs
sampling to exploit conditional structure in hidden and visible
variables), but little is said about the attributes of realized
estimates (Hinton \protect\hyperlink{ref-hinton2010practical}{2010};
Hinton, Osindero, and Teh \protect\hyperlink{ref-hinton2006fast}{2006}),
including the absence of interval estimates.

@@

While the computational concerns are inconvenient in fitting, issues due
to model flexibility are potentially more concerning. For a RBM model
with enough hidden variables, it has been shown that any distribution
for the visibles can be approximated arbitrarily well (Le Roux and
Bengio \protect\hyperlink{ref-le2008representational}{2008}; Montufar
and Ay \protect\hyperlink{ref-montufar2011refinements}{2011}; and
Montúfar, Rauh, and Ay
\protect\hyperlink{ref-montufar2011expressive}{2011}). However, for
binary data, the empirical distribution of an observed training set of
visible variables provides a highest likelihood benchmark, before any
parametric model class is even introduced and applied to obtain a
refinement of model fit. As a consequence, we find that any fully
principled fitting method based on the likelihood for a RBM with enough
hidden variables will seek to reproduce the (discrete) empirical
distribution of a training set. This aspect can be undesirable, and
perhaps even unexpected compared to most modeling scenarios, in that no
``smoothed distribution'' may result when fitting a RBM model of
sufficient size with a rigorous likelihood-based method. We are
therefore led to be skeptical that models that involve these structures
(like deep Boltzmann machines) can achieve useful prediction or
inference in a principled way without intentionally limiting the
flexibility of the fitted model.

Notions of weight-decay (penalization) and sparsity (regularization)
have been suggested in the RBM literature as practical remedies for both
over-fitting and poor mixing in the Markov chain during fitting (Hinton
\protect\hyperlink{ref-hinton2010practical}{2010}; Tieleman
\protect\hyperlink{ref-tieleman2008training}{2008}; Cho, Ilin, and Raiko
\protect\hyperlink{ref-cho2012tikhonov}{2012}). Both \(L_1\) and \(L_2\)
type penalties are mentioned to achieve weight-decay, though the
benefits of these particular forms are unknown in different situations.
However, the degree to which regularization and penalization are used by
practitioners is not clear because these concepts are not (by default) a
part of the accepted fitting methodology (Hinton
\protect\hyperlink{ref-hinton2002training}{2002}; Carreira-Perpinan and
Hinton \protect\hyperlink{ref-carreira2005contrastive}{2005}). In this
paper, we attempt to address the concerns with overfitting and poor
mixing in an alternative (and perhaps more transparent) manner, via
specification of a Bayesian prior in Section
\ref{bayesian-model-fitting}.

This paper is structured as follows. Section \ref{background} formally
defines the RBM including the joint distribution of hidden and visible
variables and explains the model's connection to deep learning.
Additionally, measures of model impropriety and methods of
quantifying/detecting it are defined. Section
\ref{explorations-of-model-properties-through-simulation} details our
explorations into the model behavior and potential propriety issues with
the RBM class. This follows with the call to action in Sculley et al.
(\protect\hyperlink{ref-sculley2018winner}{2018}) to employ small toy
examples to advance empirical understanding of deep learning models and
strive for a ``higher level of empirical rigor in the field.'' We
examine three Bayesian fitting techniques intended to avoid model
impropriety issues raised in Section \ref{model-fitting} and conclude
with a discussion in Section \ref{discussion}. Supplementary online
material provides proofs for results on RBM parameterizations and data
codings described in Section \ref{data-encoding}.

>>>>

\maketitle
\begin{abstract}
A restricted Boltzmann machine (RBM) is an undirected graphical model
constructed for discrete or continuous random variables, with two
layers, one hidden and one visible, and no conditional dependency within
a layer. In recent years, RBMs have risen to prominence due to their
connection to deep learning. By treating a hidden layer of one RBM as
the visible layer in a second RBM, a deep architecture can be created.
RBMs are thought to thereby have the ability to encode very complex and
rich structures in data, making them attractive for supervised learning.
However, the generative behavior of RBMs is largely unexplored and
typical fitting methodology does not easily allow for uncertainty
quantification in addition to point estimates. In this paper, we discuss
the relationship between RBM parameter specification in the binary case
and model properties such as degeneracy, instability and
uninterpretability. We also describe the associated difficulties that
can arise with likelihood-based inference and further discuss the
potential Bayes fitting of such (highly flexible) models, especially as
Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM
model structure.
\end{abstract}
\noindent%
{\it Keywords:}  Degeneracy, Instability, Classification, Deep Learning, Graphical Models
\vfill

@@

First, the models can be unsatisfactory as conceptualizations of how
data are generated. That is, recalling Fisher
(\protect\hyperlink{ref-fisher1922mathematical}{1922}), the aim of a
statistical model is to represent data in a compact way. Neyman and Box
further state that a model should ``provide an explanation of the
mechanism underlying the observed phenomena'' (Lehmann
\protect\hyperlink{ref-lehmann1990model}{1990}; G. E. P. Box
\protect\hyperlink{ref-box1967discrimination}{1967}). At issue,
simulation from RBMs can often produce data lacking realistic
variability so that such models may thereby fail to satisfactorily
reflect a true data generation process. Such behavior relates to model
degeneracy (or near degeneracy), which is a statistical concern in that
most data processes being modeled are realistically not degenerate in
their spectrum of potential outcomes. For example, when sampling data
from a nearly degenerate RBM, only a small number of output
possibilities receives nearly all probability, and thus a sample of
images will all be copies of the same one image (or small number of
images). An example of ten 4-pixel images simulated from a nearly
degenerate RBM model is compared to ten 4-pixel images simulated from a
non-degenerate RBM model in Figure \ref{fig:sample-models}. The
degenerate model places almost all probability on one outcome, causing
the image to be generated repeatedly, whereas the non-degenerate model
shows more potentially realistic variation.

@@

Separate of the exploration of model properties, we also investigate the
quality of estimation in fitting these models and examine an estimation
procedure that allows for uncertainty quantification rather than point
estimates alone. The fitting of RBMs can be problematic for two reasons,
the first being computational and the second concerning flexibility. As
the size of these models grows, both maximum likelihood and Bayesian
methods of fitting quickly become intractable. The literature often
suggests Markov chain Monte Carlo (MCMC) tools for approximate
maximization of likelihoods to fit these models (e.g., Gibbs sampling to
exploit conditional structure in hidden and visible variables), but
little is said about the attributes of realized estimates (Hinton
\protect\hyperlink{ref-hinton2010practical}{2010}; Hinton, Osindero, and
Teh \protect\hyperlink{ref-hinton2006fast}{2006}), including the absence
of interval estimates.

@@

As a compounding factor to the computational challenges, issues in model
fitting due to model flexibility are potentially more concerning. For a
RBM model with enough hidden variables, it has been shown that any
distribution for the visibles can be approximated arbitrarily well (Le
Roux and Bengio \protect\hyperlink{ref-le2008representational}{2008};
Montufar and Ay \protect\hyperlink{ref-montufar2011refinements}{2011};
and Montúfar, Rauh, and Ay
\protect\hyperlink{ref-montufar2011expressive}{2011}). However, for
binary data, the empirical distribution of an observed training set of
visible variables provides a highest likelihood benchmark, before any
parametric model class is even introduced and applied to obtain a
refinement of model fit. As a consequence, we find that any fully
principled fitting method based on the likelihood for a RBM with enough
hidden variables will seek to reproduce the (discrete) empirical
distribution of a training set. This aspect can be undesirable, and
perhaps even unexpected compared to most modeling scenarios, in that no
``smoothed distribution'' may result when fitting a RBM model of
sufficient size with a rigorous likelihood-based method. We are
therefore led to be skeptical that models that involve these structures
(like deep Boltzmann machines) can achieve useful prediction or
inference in a principled way without intentionally limiting the
flexibility of the fitted model.

Notions of weight-decay (penalization) and sparsity (regularization)
have been suggested in the RBM literature as practical remedies for both
over-fitting and poor mixing in the Markov chain during fitting (Hinton
\protect\hyperlink{ref-hinton2010practical}{2010}; Tieleman
\protect\hyperlink{ref-tieleman2008training}{2008}; Cho, Ilin, and Raiko
\protect\hyperlink{ref-cho2012tikhonov}{2012}). Both \(L_1\) and \(L_2\)
type penalties are mentioned to achieve weight-decay, though the
benefits of these particular forms are unknown in different situations.
However, the degree to which regularization and penalization are used by
practitioners is not clear because these concepts are not (by default) a
part of the overall model-specification-and-fitting methodology (Hinton
\protect\hyperlink{ref-hinton2002training}{2002}; Carreira-Perpinan and
Hinton \protect\hyperlink{ref-carreira2005contrastive}{2005}). In this
paper, we attempt to address the concerns with overfitting and poor
mixing in an alternative (and perhaps more transparent) manner, via
specification of a Bayesian prior in Section
\ref{bayesian-model-fitting}.

This paper is structured as follows. Section \ref{background} formally
defines the RBM including the joint distribution of hidden and visible
variables and explains the model's connection to deep learning.
Additionally, measures of model impropriety and methods of
quantifying/detecting it are defined. Section
\ref{explorations-of-model-properties-through-simulation} details our
explorations into the model behavior and potential propriety issues with
the RBM class. This aligns with the call to action in Sculley et al.
(\protect\hyperlink{ref-sculley2018winner}{2018}) to employ small toy
examples to advance empirical understanding of deep learning models and
strive for a ``higher level of empirical rigor in the field.'' We
examine three Bayesian fitting techniques intended to avoid model
impropriety issues raised in Section \ref{model-fitting} and conclude
with a discussion in Section \ref{discussion}. Supplementary online
material provides proofs for results on RBM parameterizations and data
codings described in Section \ref{data-encoding}.

### 2018-09-02 writing/paper-rbm/index.Rmd
intro wording tweaks

---
title: "Properties and Bayesian fitting of restricted Boltzmann machines"
author: 
  - name: Andee Kaplan
    university: Duke University
    department: Department of Statistical Science
  - name: Daniel Nordman
    university: Iowa State University
    department: Department of Statistics
  - name: Stephen Vardeman
    university: Iowa State University
    department: |
      | Department of Statistics and Department of Industrial and
      | Manufacturing Systems Engineering
site: bookdown::bookdown_site
output: 
  bookdown::pdf_book:
    includes:
      in_header: ../resources/latex/header.tex
    toc: no
bibliography: [../resources/latex/refs_rbm.bib]
github-repo: andeek/rbm
abstract: |
   A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, and no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning. By treating a hidden layer of one RBM as the visible layer in a second RBM, a deep architecture can be created. RBMs are thought to thereby have the ability to encode very complex and rich structures in data, making them attractive for supervised learning. However, the generative behavior of RBMs is largely unexplored and typical fitting methodology does not easily allow for uncertainty quantification in addition to point estimates. In this paper, we discuss the relationship between RBM parameter specification in the binary case and model properties such as degeneracy, instability and uninterpretability. We also describe the difficulties that arise in likelihood-based and discuss potential Bayes fitting of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
keywords: Degeneracy, Instability, Classification, Deep Learning, Graphical Models
blind: false
---

>>>>

---
title: "Properties and Bayesian fitting of restricted Boltzmann machines"
author: 
  - name: Andee Kaplan
    university: Duke University
    department: Department of Statistical Science
  - name: Daniel Nordman
    university: Iowa State University
    department: Department of Statistics
  - name: Stephen Vardeman
    university: Iowa State University
    department: |
      | Department of Statistics and Department of Industrial and
      | Manufacturing Systems Engineering
site: bookdown::bookdown_site
output: 
  bookdown::pdf_book:
    includes:
      in_header: ../resources/latex/header.tex
    toc: no
bibliography: [../resources/latex/refs_rbm.bib]
github-repo: andeek/rbm
abstract: |
   A restricted Boltzmann machine (RBM) is an undirected graphical model constructed for discrete or continuous random variables, with two layers, one hidden and one visible, and no conditional dependency within a layer. In recent years, RBMs have risen to prominence due to their connection to deep learning. By treating a hidden layer of one RBM as the visible layer in a second RBM, a deep architecture can be created. RBMs are thought to thereby have the ability to encode very complex and rich structures in data, making them attractive for supervised learning. However, the generative behavior of RBMs is largely unexplored and typical fitting methodology does not easily allow for uncertainty quantification in addition to point estimates. In this paper, we discuss the relationship between RBM parameter specification in the binary case and model properties such as degeneracy, instability and uninterpretability. We also describe the associated difficulties that can arise with likelihood-based inference and further discuss the potential Bayes fitting of such (highly flexible) models, especially as Gibbs sampling (quasi-Bayes) methods are often advocated for the RBM model structure.
keywords: Degeneracy, Instability, Classification, Deep Learning, Graphical Models
blind: false
---
